---
title: "Exercise - Answer Key"  
format:
  html:
    toc: true
    toc-title: "Exercise"  
    number-sections: false 
---
  
## Lab 1 

Load the dataset "mpg" and work through the exercises below. Note, "mpg" is included in the tidyverse package, so you will need to load the package first.

1. Calculate the **mean**, **range**, **minimum**, and **maximum** of the variable "hwy" across all models. Then, combine these statistics into one vector. (Tip: look up the RDocumentation for the functions `mean`, `range`, `min`, and `max`).

```{r} 

library(tidyverse)
data(mpg)

mean_hwy <- mean(mpg$hwy)
range_hwy <- range(mpg$hwy)
min_hwy <- min(mpg$hwy)
max_hwy <- max(mpg$hwy)
sum_stats <- c(mean_hwy, range_hwy, min_hwy, max_hwy) 
sum_stats
```

2. Since "hwy" is measured in miles per gallon, create a new variable in mpg that expresses "hwy" in litres per 100 km. 

```{r} 

temp <- 235.215 / mpg$hwy
mpg$hwy_Lp100km_1 <- temp

# Or, in one step
mpg$hwy_Lp100km_2 <- 235.215 / mpg$hwy

unique(mpg$hwy_Lp100km_1 == mpg$hwy_Lp100km_2) # sanity check
```

3. Identify the models of cars that are most fuel-efficient. Which classes of cars are least fuel-efficient? 

```{r} 

mpg$model[mpg$hwy == max(mpg$hwy)]

mpg$class[mpg$hwy_Lp100km_1 == max(mpg$hwy_Lp100km_1)]

unique(mpg$class[mpg$hwy_Lp100km_1 == max(mpg$hwy_Lp100km_1)]) # using unique function to remove duplicates
```

4. Compute the quantiles of "hwy". Can you also calculate the tertiles instead? (Tip: look up for the RDocumentation for the function `quantile`).

```{r}

quantile(mpg$hwy)

quantile(mpg$hwy, probs = seq(0, 1, 0.333)) 
quantile(mpg$hwy, c(0, 0.333, 0.666, 1)) # you can achieve the same using this
```

5. Now, based on the tertiles you calculated, assign "least efficient", "medium", and "most efficient" labels to all models. Try using both base R indexing and the function `ifelse`.

```{r}

tertile <- quantile(mpg$hwy, c(0, 0.333, 0.666, 1))

mpg$efficient_class_1 <- NA # optional, this eliminates the warning message of "Unknown or uninitialised column"
mpg$efficient_class_1[mpg$hwy < tertile[2]] <- "least efficient"
mpg$efficient_class_1[mpg$hwy >= tertile[2] & mpg$hwy < tertile[3]] <- "medium"
mpg$efficient_class_1[mpg$hwy >= tertile[3]] <- "most efficient"

# OR
mpg$efficient_class_2 <- NA # given how ifelse functions are specified below, this step becomes necessary
mpg$efficient_class_2 <- ifelse(mpg$hwy < tertile[2], "least efficient", mpg$efficient_class_2)
mpg$efficient_class_2 <- ifelse(mpg$hwy >= tertile[2] & mpg$hwy < tertile[3], "medium", mpg$efficient_class_2)
mpg$efficient_class_2 <- ifelse(mpg$hwy >= tertile[3], "most efficient", mpg$efficient_class_2)

unique(mpg$efficient_class_1 == mpg$efficient_class_2) # sanity check
```
  
## Lab 2 

### Data Manipulation using dplyr

#### Part A {-}
 
Continue from the farm business dataset "farm" used in Lab 2, work through the exercises below.

  1. Create a new variable called **size3** that meets the following criteria: 
  
      i) size3 = “small” if size <= 200
    
      ii) size3 = “medium” if 200 < size <= 600
    
      iii) size3 = “big” if size > 600
    
      Finally, convert size3 to a factor variable that is ordinal from "small" to "big".  
  
```{r}
#| warning: false
#| message: false

name <- c("Henry", "Larry", "Alex", "Gaby", "Amy", "Ruby")
sex <- c("male", "male", "male", "female", "female", "female")
age <- c(43, 60, 25, 50, 28, 58) 
type <- c("crop", "livestock", "urban", "dairy", "crop", "livestock") 
size <- c(550, 800, 10, 600, 1000, 700) 
return <- c(40, 90, 50, 90, 90, 95)

farm <- data.frame(name, age, sex, type, size, return)
 
farm <- farm %>% mutate(size3 = ifelse(size <= 200, "small", NA)) %>%
                 mutate(size3 = ifelse(size > 200 & size <= 600, "medium", size3)) %>%
                 mutate(size3 = ifelse(size > 600, "big", size3)) %>%
                 mutate(size3 = factor(size3, levels = c("small", "medium", "big")))

farm
farm$size3  

```
 
  2. Generate the following summary statistics, for each type of the farms:
  
      i) the sum of all returns, called **tot.return**
    
      ii) the average returns, called **avg.return**
    
      Finally, rearrange the data based on the value of avg.return, in the descending order.
  
```{r}
#| warning: false
#| message: false

farm.sum <- farm %>% group_by(type) %>%
                     summarize(tot.return = sum(return), 
                               avg.return = mean(return))

farm.sum <- arrange(farm.sum, desc(avg.return))
farm.sum

```

#### Part B {-}
  
Import data "mpg" and work through the coding below.
   
  1. Drop the variables **displ**, **drv** and **fl**, then exclude cars that were manufactured by **Hyundai** and **Pontiac** .

```{r}
#| warning: false
#| message: false

data(mpg)     # you may need to library(tidyverse) first

mpg <- mpg %>% select(-displ, -drv, -fl) %>% 
               filter(manufacturer != "hyundai" & manufacturer != "pontiac")

head(mpg)      

unique(mpg$manufacturer)

```

  2. Continue from your dataframe above and generate the summary statistics, for each manufacturer, model, and year:
  
      i) the average of "cty", called **avg.cty**
    
      ii) the average of "hwy", called **avg.hwy**
    
      iii) the total number of cars produced, called **tot.cars**
  
```{r}
#| warning: false
#| message: false

mpg.sum <- mpg %>% group_by(manufacturer, model, year) %>%
                   summarize(avg.cty = mean(cty),
                             avg.hwy = mean(hwy),
                             tot.cars = n())

head(mpg.sum)

``` 

### Data Visualization using ggplot

#### Part A {-}

Import data "mpg" and work through the exercises below.

  1. Create a scatterplot that shows all of the following:
  
      i) use "cty" on the x-axis and "hwy" on the y-axis
        
      ii) assign different colors for "class" and different shapes for "year".  
        (tip: since continuous variables cannot be mapped to `shape`, you will need to convert "year" to a factor using `factor()`)
       
      iii) fit **only one trend line** for all points 
  
```{r}
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 6

data(mpg)     # library(tidyverse)

ggplot(mpg) +
  geom_point(aes(x= cty, y = hwy, color = class, shape = factor(year))) +
  geom_smooth(aes(x= cty, y = hwy))

```

  2. Create a scatterplot that shows all of the following:
  
      i) use "displ" on the x-axis and "hwy" on the y-axis
        
      ii) partition the graph into a grid by the combination of "drv" and "fl", but only include "fl" that equals "p" or "r"
        
      iii) fit a trend line in each graph within the grid
    
      (Tip: you can build from the example "p23" in Lab 2.) 
  
```{r}
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 6
 
mpg.mod <- mpg %>% filter(fl == "p" | fl == "r")

ggplot(mpg.mod, aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth() + 
  facet_grid(drv ~ fl)

```

  3. Create a bar chart that shows the average "hwy" of all cars produced in 2008 by each of the following manufacturers:  
  **Audi**, **Hyundai**, **Nissan** and **Volkswagen**. Assign a different color to each manufacturer. 

```{r}
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 6

mpg.sum <- mpg %>% filter(manufacturer == "audi" | manufacturer == "hyundai" |
                          manufacturer == "nissan" | manufacturer == "volkswagen") %>%
                   filter(year == 2008) %>% 
                   group_by(manufacturer) %>%
                   summarize(mean_hwy = mean(hwy))

ggplot(mpg.sum, aes(x = manufacturer, y = mean_hwy, fill = manufacturer)) + geom_col()

```

#### Part B {-}

Import the data "gapminder" and try to reproduce the graph below.

```{r}
#| warning: false
#| message: false
#| fig-width: 12
#| fig-height: 6
 
library(gapminder)

ggplot(gapminder, aes(x = gdpPercap, y = lifeExp, color = continent)) +
  geom_point(shape = 17, size = 1.5) + 
  geom_smooth() + 
  facet_grid(continent ~ year, scales = "free") +
  
  labs(
    title = "Relation between GDP per capita and Life Expectancy by Continent from 1952 to 2007",
    x = "GDP per capita", y = "Life Expectancy", color = "Legend"
  ) +
  
  theme(
    axis.text.x = element_blank(), 
    title = element_text(face = "bold")
  )
 
```

## Lab 3 

Continue from your multiple linear regression "mod2":

1. Calculate the R-squared using $y_i$ and $\hat{y}_i$, that is
  $$
  R^2 = corr(y_i, \hat{y}_i)^2
  $$
  (tip: you can retrieve the fitted values $\hat{y}_i$ via `mod2$fitted.values`.)
  
```{r}

wage <- read.csv(file = "D:/Dropbox/Teaching/AREC513/Fall2025/materials/data/ch2_data_csv/wages1.csv") 
mod2 <- lm(wage ~ male + school + exper, data = wage)
sum_mod2 <- summary(mod2)

y     <- wage$wage
yhat  <- mod2$fitted.values
R2_via_corr <- cor(y, yhat)^2
R2_via_corr

```

2. Construct the confidence interval for $\hat{\beta}_{male}$, with the significance level of 10%.

```{r}

ci_male_90 <- confint(mod2, parm = "male", level = 0.90)
ci_male_90

```

{{< pagebreak >}}

3. Test $H_0:\beta_{exper} = 0.15$,  with the significance level of 10%. 

```{r}

beta_exper <- sum_mod2$coefficients["exper","Estimate"]
se_exper   <- sum_mod2$coefficients["exper","Std. Error"]
df         <- mod2$df.residual
t_stat     <- (beta_exper - 0.15) / se_exper

t_crit <-  qt(1 - 0.10 / 2, df)
t_stat >= t_crit   # gives False, hence do not reject H0

```

4. Test $H_0:\beta_{school} + \beta_{exper} = 0.75$,  with the significance level of 1%.

```{r}
#| message: false

library(car)
linearHypothesis(mod2, c("school + exper = 0.75"))   # do not reject H0, given the F stats or p-value

```

5. Predict the wage per hour for a male worker with 18 years of schooling and 0 years of experience.

```{r} 

pred_data <- data.frame(male = 1, school = 18, exper = 0)

predict(mod2, newdata = pred_data)

```

## Lab 4

```{r}
#| echo: false
#| message: false   

library(stargazer)   
library(lmtest)  
library(car)         
library(estimatr)  
library(effects)    
library(AER)
```
 
1. Estimate a similar regression as "eq.a" using the data "males" but **without the intercept**, report the regression results.  
  (tip: check the RDocumentation of `lm()` to see how to estimate a model without the intercept.)
```{r}
#| warning: false

setwd("D:/Dropbox/Teaching/AREC513/Fall2025/materials/data")
males <- read.csv(file = "ch3_data_csv/males1987_updated.csv")
names(males) <- tolower(names(males))

eq.1 <- logwage ~ school + exper + union + mar + black + hisp - 1

mod.1 <- lm(eq.1, data = males)
summary(mod.1)
```

2. Why does the Sargan Test report NA values for object "diag_iv"?

```
Because Sargan Test is for the over-identified case, where the number of IVs is larger than the number of endogeneous variables.
```

3. Load in the data "school":

```{r} 
#| warning: false
setwd("D:/Dropbox/Teaching/AREC513/Fall2025/materials/data")
school <- read.csv(file = "ch5_data_csv/schooling.csv") 

school <- school[, c("black", "smsa76", "south76", "age76", "ed76", "daded", "momed", "wage76")]
```

   (i) Estimate the regression
   $$
   \log(wage76_i)=\beta_1 ed76_i + \beta_2 age76_i + \beta_3 black_i + \beta_4 smsa76_i + \beta_5south76_i + \epsilon_i
   $$
```{r}
eq.2 <- log(wage76) ~ ed76 + age76 + black + smsa76 + south76
mod.2 <- lm(eq.2, data = school)
summary(mod.2) 
```
   
   (ii) Suppose we suspect there's a non-linear relationship between wage and one's age, how would you change your regression?
   
```{r}
eq.3 <- log(wage76) ~ ed76 + age76 + I(age76 ^ 2) + black + smsa76 + south76
mod.3 <- lm(eq.3, data = school)
summary(mod.3) 
```
  
   (iii) Check whether your model in (ii) suffers from a multicollinearity or heteroskedasticity problem.
   
```{r}
vif(mod.3)     # only multicollinearity between age and age^2, but it's expected
   
bptest(mod.3)  # only weak evidence of heteroskedasticity at the 10% significance level
```
  
   (iv) Now perform a subsample analysis based on whether the individual was living in the South in 1976. 
```{r}
mod.4 <- lm(eq.3, data = school, subset = (south76 == 1)) 
mod.5 <- lm(eq.3, data = school, subset = (south76 == 0))
   
summary(mod.4) 
summary(mod.5)  
```
 
## Lab 5

Use the following dataset on 10 farmers that:

- a potential profit $profit^1$ is realized if the new technology is adopted, otherwise $profit^0$

- the counterfactual is observable

Answer Question 1 and 2.

```{r}
farmer_id <- 1 : 10
profit_state1 <- c(120, 90, 90, 150, 80, 160, 60, 95, 100, 75)    # adoption of the technology
profit_state0 <- c(100, 95, 85, 110, 90, 100, 70, 90, 80, 100)    # no adoption
```
 
1. Suppose the farmers will only adopt the technology if it is profitable, generate the dummy variable "d_adopt".

```{r}
df_farm <- data.frame(farmer_id, profit_state1, profit_state0)
df_farm <- df_farm %>% mutate(d_adopt = ifelse(profit_state1 > profit_state0, 1, 0)) 
```

2. Calculate $ATE$, $ATT$, $ATU$, and show that $ATE$ is the weighted average of the other two terms.

```{r} 
ATE <- mean(df_farm$profit_state1) - mean(df_farm$profit_state0)

treated_farm <- df_farm %>% filter(d_adopt == 1)  
ATT <- mean(treated_farm$profit_state1) - mean(treated_farm$profit_state0)

untreated_farm <- df_farm %>% filter(d_adopt == 0)  
ATU <- mean(untreated_farm$profit_state1) - mean(untreated_farm$profit_state0)

c(ATE = ATE, ATT = ATT, ATU = ATU)
```

Now, imagine a more realistic situation where the counterfactual is not observable, and only the realized profits and whether the farms adopted the technology can be observed. 

Create this new data frame and answer Question 3 through 5.

3. Calculate the $ODM$ and verify that the regression results correspond to it.

```{r}
ODM <- mean(df_farm$profit_state1[df_farm$d_adopt == 1]) - mean(df_farm$profit_state0[df_farm$d_adopt == 0])
ODM

df_farm <- df_farm %>% mutate(profit_obs = ifelse(d_adopt == 1, profit_state1, profit_state0))

reg_farm <- lm(profit_obs ~ d_adopt, data = df_farm)
summary(reg_farm) 
```

4. Calculate the selection bias and the heterogeneous treatment effects bias.

```{r}

treated_state0_profit   <- mean(treated_farm$profit_state0)
untreated_state0_profit <- mean(untreated_farm$profit_state0) 

selection_bias <- treated_state0_profit - untreated_state0_profit
selection_bias

# ATT + selection_bias == ODM => TRUE
```

5. Simulate a sample of 10,000 observations by repeatedly using the same 10 observations on the farmers, with only 6 out of 10 farmers adopt the technology randomly each time. Show that when estimating a regression, the coefficient of "d_adopt" approaches the true ATE.

```{r}
set.seed(1)

df_base <- function() {
  data.frame(profit_state1, profit_state0) %>% 
    mutate(
      d_adopt = ifelse(row_number() %in% sample(10, 6), 1, 0),
      profit_obs = ifelse(d_adopt == 1, profit_state1, profit_state0)
    )  
}

df_sim10000 <- bind_rows(replicate(1000, df_base(), simplify = FALSE))

reg_sim <- lm(profit_obs ~ d_adopt, data = df_sim10000)
summary(reg_sim) 
```

