---
title: "Lab 3: Linear Regression and Statistical Inference"
author: "Feng Qiu, Liyuan Xuan"
date: "Sept 29, 2025"
format:
  html:
    toc: true
    toc-title: "Section"
    df-print: default
---

## Preparation

### Installing and Loading Packages

```{r}
# load packages that might be used in this section

# install.packages(c("broom", "knitr", "car"))  # use this line to install the necessary packages

library(broom)  # to extract summary information into a tidy data frame
library(knitr)  # for better visualization of tables
library(car)    # for hypothesis testing

```

### Read in External Data

```{r}
#| warning: false

# read in data

setwd("D:/Dropbox/Teaching/AREC513/Fall2025/materials/data/ch2_data_csv")
wage <- read.csv(file = "wages1.csv") # make sure you are in the correct working directory

# alternatively, you can do 
# wage <- read.csv(file = "YOUR_FOLDER_PATH/wages1.csv") 

head(wage)

```

## Summary Statistics 

```{r}
#| message: false

sum_table_1 <- summary(wage)
sum_table_1

sum_table_2 <- tidy(wage) 
sum_table_2

sum_table_3 <- sum_table_2[, c(1, 2, 3, 4, 8, 9, 10, 11)]
# or, try sum_table_3 <- sum_table_2[, c("column", "n", "mean", "sd", "min", "max", "range", "skew")]
sum_table_3

kable(sum_table_3, caption = "Descriptive Statistics", digits = 2)

```

To acquire summary statistics by group, you can do

```{r}
# wage rate by gender
tapply(wage$wage, wage$male, summary)
```

## Ordinary Least Squares (OLS) Regression

### Simple Linear Regression

```{r}
# estimate a simple OLS
mod1 <- lm(wage ~ male, data = wage)
sum_mod1 <- summary(mod1)
print(sum_mod1) 

# return names of "mod1", so you can extract the output you need via "$"
names(mod1)
mod1$coefficients

names(sum_mod1)
sum_mod1$coefficients

```

### Multiple Linear Regression

```{r}
# estimate a multiple OLS
mod2 <- lm(wage ~ male + school + exper, data = wage)
sum_mod2 <- summary(mod2)
print(sum_mod2, signif = F)  # significance asterisk removed 

# extract the goodness of fit
gof <- c(r.squared = sum_mod2$r.squared, adj.r = sum_mod2$adj.r.squared)
gof

```

## Results of OLS Regressions

### Variance-Covariance Matrix

Recall a typical variance-covariance matrix, showing variance along the diagonal and covariance off the diagonal.

$$
\text{Var}[X] =
\begin{bmatrix}
     \text{Var}(X_1) & \text{Cov}(X_1, X_2) & \text{Cov}(X_1, X_3) & \cdots & \text{Cov}(X_1, X_n) \\
         \vdots      &    \text{Var}(X_2)   & \text{Cov}(X_2, X_3) & \cdots & \text{Cov}(X_2, X_n) \\
         \vdots      &       \vdots         &     \text{Var}(X_3)  & \cdots & \text{Cov}(X_3, X_n) \\
         \vdots      &       \vdots         &         \vdots       & \ddots &        \vdots        \\
\text{Cov}(X_n, X_1) & \text{Cov}(X_n, X_2) & \text{Cov}(X_n, X_3) & \cdots &     \text{Var}(X_n)
\end{bmatrix}
$$

```{r}
# acquire the variance-covariance matrix
vcv2 <- vcov(mod2)
vcv2

var_male <- vcv2[2, 2]
var_male
 
se_male <- sqrt(var_male) # check the standard error of beta_male
se_male

cov_male_school <- vcv2[2, 3]  # or vcv2[3, 2]
cov_male_school 

```

### Confidence Interval

Recall the formula for the confidence interval:

$$
CI_{1-\alpha}=[\hat{\beta_i} - t_{1-\alpha/2,df} \times SE(\hat{\beta_i}), \hat{\beta_i} + t_{1-\alpha/2,df} \times SE(\hat{\beta_i})]
$$

#### Calculate Confidence Interval using `confint()`
  
```{r}
ci <- confint(mod2) 
ci
```

:::callout-note
`confint()` has to be used on a fitted model object, such as "mod2" (rather than "sum_mod2"). For details, check `?confint()`.
:::


#### Calculate Confidence Interval Manually

```{r}
beta <- sum_mod2$coefficients[, 1]
se_beta <- sum_mod2$coefficients[, 2]
  
deg_f <- mod2$df.residual
alpha <- 0.05
t_crit <-  qt(1 - alpha / 2, deg_f)
  
lower_bound <- beta - t_crit * se_beta
upper_bound <- beta + t_crit * se_beta
  
ci_manual <- matrix(c(lower_bound, upper_bound), nrow = 4, ncol = 2,
                     dimnames = list(c(names(lower_bound)), # assigning row names to the new matrix
                                     c(colnames(ci)))       # assigning col names 
                    )
ci_manual
```

:::callout-tip
As we saw earlier, `qt(p, df)` is used to extract critical values for the t distribution, the "q" here means quantile function. Explore other related functions for the t distribution as, `dt(x, df)` for density, `pt(q, df)` for distribution function, and `rt(n, df)` for random generation.

Similarly:

- For the F distribution, `df(x, df1, df2)`, `pf(q, df1, df2)`, `qf(p, df1, df2)`, and `rf(n, df1, df2)`.

- For the $\chi^2$ distribution, `dchisq(x, df)`, `pchisq(q, df)`, `qchisq(p, df)`, and `rchisq(n, df)`.

- For the normal distribution, `dnorm(x)`, `pnorm(q)`, `qnorm(p)`, and `rnorm(n)`.
:::


### Hypothesis Testing

Recall the general process of hypothesis testing (two-tailed) for coefficients, for example, testing if $\beta_i$ equals a certain constant $c$

$$
H_0:\beta_i = c
$$
$$
H_1:\beta_i \neq c
$$
The calculation of the test statistic is
$$
t = \frac{\hat{\beta_i}-c}{SE(\hat{\beta_i})}
$$
The test statistic is then compared with the critical values of the t distribution $t^*_{\alpha/2, df}$, at specified significance level $\alpha$ and degree of freedom.

- $H_0$ is rejected, if $|t| \geq t^*_{\alpha/2, df}$

- $H_0$ is not rejected, if $|t| < t^*_{\alpha/2, df}$

Suppose we wish to test if $H_0:\beta_{school} = 0.5$ against $H_1:\beta_{school} \neq 0.5$ at 5% significance level: 

#### Calculate t Statistics Manually

```{r}
beta_school <- sum_mod2$coefficients[3, 1]
se_school <- sum_mod2$coefficients[3, 2]
t_stat <- (beta_school - 0.5) / se_school

t_stat >= t_crit 
```

#### Hypothesis Testing using `linearHypothesis()`

Enabled by the package "car", `linearHypothesis()` can also be used to hypothesis testing. 

```{r}
linearHypothesis(mod2, "school = 0.5")  # check the critical value for F stat using qf(1 - alpha, 3291, 3290) 
```

:::callout-important
Since `linearHypothesis()` performs an F test, it can also be used for testing on multiple variables. Try `linearHypothesis(mod2, "school = male")`. 
:::

##  Prediction with Linear Regression

Given our fitted model "mod2", we can now use it to perform prediction. Suppose we wish to predict the wage per hour for

1. A male and a female worker with 16 years of schooling

2. Other variables (i.e. "exper") are kept at the average level

```{r} 
pred_data <- data.frame(male = c(female = 0, male = 1), 
                        school = 16, 
                        exper = mean(wage$exper)
                        )
pred_data

yhat <- predict(mod2, pred_data)
yhat

# visualizing the prediction
hist(wage$wage, xlab = "Wage", main = "Histogram of Wage per Hour")
abline(v = yhat, col = c("red", "blue"), lwd = 3)

```
 
## Exercise

Continue from your multiple linear regression "mod2":

1. Calculate the R-squared using $y_i$ and $\hat{y}_i$, that is
  $$
  R^2 = corr(y_i, \hat{y}_i)^2
  $$
  (tip: you can retrieve the fitted values $\hat{y}_i$ via `mod2$fitted.values`.)

2. Construct the confidence interval for $\hat{\beta}_{male}$, with the significance level of 10%.

3. Test $H_0:\beta_{exper} = 0.15$,  with the significance level of 10%. 

4. Test $H_0:\beta_{school} + \beta_{exper} = 0.75$,  with the significance level of 1%. 

5. Predict the wage per hour for a male worker with 18 years of schooling and 0 years of experience.